---
title: 'Project 2: Modelling daily cycling demand in Edinburgh'
author: 'Group members: Alice Miller, Saba Sitter, Jake Saunders'
date: 'Word count: [XXX words]'
output:
  pdf_document:
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{amsmath}
- \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,    # This hides all code from the main body 
  message   = FALSE, 
  warning   = FALSE
)

source("code.R") # Runs all models and creates tables/plots in the background
```

# Introduction

The "cycle daily" dataset  consists of approximately 2,180 daily observations ranging from 2020 to 2025, sourced from Edinburgh's network of automatic cycling counters, along with local weather and temporal data. This data comes from sensors at specific locations across the city, so it's possible not every cyclist is accounted for. Similarly, some cyclists may pass through multiple sensors in a day and be observed more than once. However, they are a good indicator of how cycling demand varies across different days and weather patterns. Edinburgh relies heavily on public transport for its residents and tourists to navigate the city. For future funding allocation and planning, it is important to know how many cyclists are on the road in order to evaluate whether the infrastructure (cycle lanes, crossings etc) and investment in schemes (eg rentable Voi bikes) need to adapt to meet demand. The modelling goal is to develop a sequence of increasingly complex linear regression models to isolate how temporal and weather variables, such as the day of the week and the temperature, influence cycling demand. Ultimately, the goal is to extrapolate cycling demand so that Edinburgh City Council and Transport Scotland can optimise infrastructure investment over the next decade to align with their active travel priorities.


# Exploratory data analysis

To begin investigating the impact of weather on the number of daily cyclists, we begun with some exploratory data analysis. This involved converting month to an ordered factor, redefining the day of week predictor with explicit levels. Additionally, we created a new variable 'trend' which symbolises the number of days since the first observation (01/01/2020), along wtih binary indicators 'is_covid' and 'is_holiday' which represent the UK Covid lockdown (23/03/2020 - 31/03/2021) and the yearly festive period (24th - 31st December) respectively. 

We initially found the summary statistics for the number of daily cyclists and the weather data.

```{r summary stats exploratory table, echo=FALSE}

kable(summary_stats, 
             digits = 2, 
             caption = "Summary Statistics for count, temp mean, temp min, temp max.")%>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE)
```

We can see from table 1 that the number of daily cyclists varies from $1,264$ to $43,864$, a variation of $\approx 3370 \%$. Intuitively, one would think that a warmer average weather would lead to more cyclists. 

We then plotted a time series of the daily cyclist count over the period of observations. 

```{r time series, fig.cap="Time series of daily cyclist counts 2020-2025.",fig.height = 3, fig.width = 5}
plot_timeseries
```


We can see from Figure 1 that there are yearly fluctuations, with peaks and troughs occurring seasonally. Overall, the cyclist count seems to be decreasing over time. There is a visible peak on and surrounding the 10th of August 2022, which we investigated. We found that there was a London-Edinburgh-London bicycle event taking place over this period with over 2000 riders$^{1}$. Supposing each cyclist passed through multiple sensors over the course of the days, this likely explains the inflated numbers. 

We then created boxplots of the daily cyclist count by month and by the day of week to observe whether they had any significant impact on the number of daily cyclists. 


```{r Boxplots of cyclist count by month and day of week, fig.cap="Boxplots of cyclist count by month and day of week.", fig.height = 5, fig.width = 7}
grid.arrange(plot_month,plot_dow, nrow = 2)
```


We can see from Figure 2 that the colder months (October - February) have slightly lower cyclist counts, with the other months remaining relatively constant. There are a lot of outliers, especially in Spring. Observations with unusually high cyclist counts could be attributed to special events such as races, bank holidays, and exceptionally sunny days. Outside of winter, the month doesn't appear to have a tremendous impact on the median daily cyclist count. The second boxplot in Figure 2 indicates that there tends to be less cyclists on Mondays and Sundays, with the other days appearing relatively constant. 

We then plotted the daily cyclist count against the mean temperature, with the red and blue lines indicating regression and quadratic relationship respectively as seen in Figure 3. 

```{r scatter plot of count vs temp_mean, fig.cap="Scatter plot of Daily cyclist count vs mean temperature.", fig.height = 3, fig.width = 5}
plot_temp_scatter
```

There is a clear positive correlation between temperature mean and daily cyclist count, which is to be expected. There are significantly less observations for the days with the lowest and highest mean temperatures, so there is not a clear trend that we can see for those days. There is a large variation in cyclist counts for days with the same temperature, so it is not feasible to predict counts based off of mean temperature alone.  

We noticed that the days with the lowest counts fell between Christmas Eve and New Years Eve, presumably due to less commuters on the road. We created a graph to illustrate this over the course of the 5 years. 

```{r plot december, fig.cap="Line Graph of average daily cyclists across the 5 years in decemeber.", fig.height = 3, fig.width = 5}
plot_december

```

We can see from Figure 4 that there is a substantial decrease in cyclist numbers between the specified days. This led us to add whether a day falls between the 24th and 31st of December in our final model as 'is_holiday'.

The data spans the Covid-19 pandemic, so we added a binary predictor 'is_covid' into the dataframe to indicate whether a day falls within the UK mandated lockdown, which we later investigated.

# Baseline model 

We fit an initial linear model, M0, that uses the numeric variables mean temperature, weekend, and month to predict the daily cyclist count. 

The equation for M0 is:

$$\text{count}_i = \beta_0 + \beta_1 \text{temp\_mean}_i + \beta_2 \text{weekend}_i + \beta_3 \text{month}_i + \epsilon_i$$
$$\epsilon_i \sim N(0, \sigma^2)$$

The summary statistics for M0

```{r summary for m0} 

coef_matrix <- summary(m0)$coefficients
coef_matrix <- coef_matrix[, -4]
kable(coef_matrix, 
      digits = 2, 
      caption = "Summary coefficients for model M0") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) 
                
```

For every 1 degree increase in the mean daily temperature, M0 predicts an increase of 247 cyclists per day, and predicts a 1247 decrease in cyclist count on weekends compared to weekdays. Currently, the model suggests that there is a decrease of 247 cyclists for each month passing. This is because M0 uses month as a numeric predictor and therefore it ignores the fact that cyclist counts generally increase in summer and fall in winter.

We then assessed the diagnostic plots for M0.

```{r r vs f and qq plots m0, fig.cap="Diagnostic plots for M0.", fig.height = 4, fig.width = 6}
grid.arrange(p_m0_1,p_m0_2, p_m0_3,p_m0_4, nrow = 2, ncol = 2)
```

The linear assumption seems to be validated as we can see from the Scale-Location plot in Figure 6, and there don't appear to be any points with unreasonable leverage. The Residuals vs Fitted plot in Figure 5 shows a mean variance of 0, independent of the fitted values. The residuals show a non-linear pattern, with an arguable 'n' shape above the mean line, indicating there is an underlying non-linear structure yet to be included in our model. The Q-Q plot in Figure 5 demonstrates right skewing and heavy tail in the positive quantiles, indicating that M0 is particularly bad at predicting high count observations. Therefore, the normality assumption for linear regression is violated. 

# Proposed Models
## Model 1: Linear temperature with trend and factor variables

We created a further model, M1, which adds in the trend variable as a predictor to capture the drift in cycling levels over 2020-2025, and interprets day of week and month as factor variables rather than numeric variables.

Our proposed equation for M1 (identifiable form) is:
$$\text{count}_i = \beta_0 + \beta_1 \text{temp\_mean}_i + \beta_2 \text{trend}_i + \sum_{j=2}^{12} \gamma_j \mathbb{I}(\text{month}_i = j) + \sum_{k=2}^{7} \delta_k \mathbb{I}(\text{dow}_i = k) + \epsilon_i$$

$$\epsilon_i \sim N(0, \sigma^2)$$

The prediction for day 1 (Monday) and month 1 (Jan) are represented by the intercept term $\beta_0$
Every other month or day is measured as the expected difference in count from the baseline. M1 is an improvement over M0 because M1 takes into account the impact of different weekdays rather than simplistic weekend or non weekend effects. Changing 'month' to a factor variable allows for building a seasonal profile rather than assuming the month index (1-12) is linear with the count.
We omit the weekend term and numerical version of 'month' to avoid collinearity with their 'factor' versions and because the covariate 'dow' is collinear with 'weekend'.


```{r side by side comparison of M0 and M1, fig.cap="Diagnostic Plots for M1", fig.width = 6, fig.height= 4}
grid.arrange(p_m1_1,p_m1_2,p_m1_3,p_m1_4, nrow = 2, ncol = 2)
```

We can see from the Residuals vs Fitted Values plot in Figure 6 that, despite the 'n' pattern about the mean being slightly less pronounced, the mean residual line follows a slight sigmoid shape, suggesting the linear assumption is still violated for M1. The Q-Q plot looks slightly more aligned to the normal line at the lower tail in comparison to M1, suggesting that the prediction for the days with lower cyclist counts has improved. There is now however a point with very large leverage as seen in the Residuals vs Leverage plot. The linear assumption appears to be slightly more justified than with M0, as we can see from the Scale-Location plot.

## Model M2: Quadratic temperature 

Our proposed equation for M2 (identifiable form) is:
$$\text{count}_i = \beta_0 + \beta_1 \text{temp\_mean}_i + \beta_2 \text{temp\_mean}^2_i + \beta_3 \text{trend}_i + \sum_{j=2}^{12} \gamma_j \mathbb{I}(\text{month}_i = j) + \sum_{k=2}^{7} \delta_k \mathbb{I}(\text{dow}_i = k) + \epsilon_i$$
$$\epsilon_i \sim N(0, \sigma^2)$$

We can infer from Figure 3 that there is an underlying non-linear relationship between the mean temperature and the daily cyclist count. Hence, replacing the linear 'mean temperature' term with a nonlinear alternative term, the underlying structure may be better explained by the model. We can evaluate this using the following diagnostic plots for M2.

```{r 1st_Diagnostics_of_M1_M2, fig.cap="Diagnostic Plots for M2", fig.height = 4, fig.width = 6}
grid.arrange(p_m2_1, p_m2_2, p_m2_3, p_m2_4, nrow = 2, ncol = 2)
```     

We can see from Figure 7 that the linear assumption still appears to be justified and there is still a point with massive influence. There is not much difference from M1 in the diagnostic plots, suggesting that the explanatory power of M2 is not improved upon M1.


## Model M3: Logarithmic count with nonlinear temperature

The expression is: 
$$\log(\text{count}_i + 1) = \beta_0 + \beta_1 \text{temp\_mean}_i + \beta_2 \text{temp\_mean}_i^2 + \beta_3 \text{trend}_i + \sum_{j=2}^{12} \gamma_j \mathbb{I}(\text{month}_i = j) + \sum_{k=2}^{7} \delta_k \mathbb{I}(\text{dow}_i = k) +
$$$$
\beta_4 \text{is\_covid}_i + \beta_5 \text{is\_holiday}_i + \epsilon_i$$
$$\epsilon_i \sim N(0, \sigma^2)$$

To mediate the heavy right tailed Q-Q plots seen in the previous models, we apply the log 
transform to the count response. We added in binary predictor variables - 'is_holiday' and 'is_covid' - introduced and defined during exploratory data analysis. We made the following graphs to justify the validity of adding these predictor variables to M3.


```{r Boxplot of holiday, fig.cap="Boxplot of 'is_holiday' variable (Dec 24th - 31st)", fig.height = 3, fig.width = 5}
plot_holiday
```

We can see from Figure 8 that the median daily cyclist count is significantly lower during the specified festive period. The spread (IQR) of counts is also very small during the festive period, indicating that there are consistently much lower counts during this time. This suggests that it is a valid indicator to include. 

```{r Impact of adding covid, fig.cap="Overall trend during lockdown and over the whole period", fig.width = 6, fig.height = 3}
plot_covid

```

We can see from Figure 9 that during the lockdown period, the decline is more pronounced than for the whole time period that the data spans. This indicates that there is an observable effect of Covid on the cyclist count, and that it is justified in adding it into M3. 

We then evaluated the assumptions of M3 through looking at the following diagnostic plots. 

```{r Diagnostic plots M3, fig.cap="Diagnostic Plots for M3", fig.height = 4, fig.width = 6}
grid.arrange(p_m3_1, p_m3_2, p_m3_3, p_m3_4, nrow = 2, ncol = 2)
```

The difference in the linear assumption is negligible compared to the previous models. The Q-Q plot suggests that the normality assumption is more justified than in the previous models. The scale-location plot suggests that M3 displays greater heteroscedasticity, meaning the constant variance assumption is violated more than in M2. There is still a point of high influence as shown in the Residuals vs Leverage plot. 

Our final model is M2. We created a coefficient table for it as follows.

```{r summary for m2} 

cycle_daily_df$month <- factor(cycle_daily_df$month, ordered = FALSE)
cycle_daily_df$dow <- factor(cycle_daily_df$dow, ordered = FALSE)

m2new <- lm(count ~ temp_mean + I(temp_mean^2)  + trend + 
           factor(month) + factor(dow), data = cycle_daily_df )
m2_table <- as.data.frame(summary(m2new)$coefficients)
colnames(m2_table) <- c("estimate", "std Error", "t value", "p value")

kable(m2_table, 
      digits = 4, 
      caption = "Summary for Model M2") 
                
```

We can see from Table 3 that the estimated daily cyclist count when the mean temperature is 0, and it is a Monday during January is 6990. The coefficient estimates represent the estimated difference from the baseline for each predictor. There is a positive coefficient for the mean temperature variable, and a negative one for the quadratic mean temperature variable. This indicates that the cyclist count rises generally with the temperature, yet is lower at the more extreme high and low mean temperatures. All of the month coefficients are positive except December, which indicates that January sees the second-lowest daily cyclist counts. The day of week coefficients are all positive, which makes sense as we saw during exploratory data analysis that Monday has the lowest average cyclist count. The p-values are all very low (under 0.03) except from the month October and the day Sunday. Therefore, we can conclude that the predictors are significant at a 97% significance level. 

# Evaluation
## Statistical Performance

We found the following error scores for each model: root mean squared, mean absolute error, Dawid-Sebastiani score, and interval score. For each year in 2020-2025, we found the scores when that year was removed from the training data and used as the test data, and then found the average score for each model. 


```{r Table 1 LOYO CV scores for M0-M3}

kable(table1_final, 
             digits = 2, 
             caption = "Leave-one-year-out CV scores")  %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) 
```

Since these scores are representative of various errors in the models, lower scores are preferable.
We can see from Table 4 that M2 has the lowest score for RMSE, which indicates that M2 fits the dataset the best. M2 also has the lowest score for MAE, suggesting it is less sensitive to extreme outliers, and IS, demonstrating that M2 produces narrow, specific intervals that include the true count most of the time. M0 has a lower DS than M2 by 0.02, suggesting it is slightly better at predicting the mean and variance of the daily cyclist count. All of the DS scores are very close to each other, and we already established that M0 is the worst model as it doesn't take any quadratic effects of temperature, or the fact that there are factor variables, into account. 

This was the basis for us choosing M2 as our final model. 

We then created a table showing the RMSE and DS for each month using M2.

```{r Table 2 RMSE and DS by month}

kable(table2_final, 
             digits = 2, 
             caption = "Table of the RMSE and DS by months")  %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) 
```

## Practical Performance

### Temperature effects and nonlinearlity

The following table fits M2 three times, using temp mean, temp min and temp max in turn as the basis for the quadratic temperature term, and reports the scores. 

```{r M2 fit scores with temp_mean, temp_min, and temp_max}

kable(m2_comparison, 
             digits = 2,
             caption = "Model M2 fitted with daily temp mean, temp min, temp max.") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) 
```

The model fit with the minimum daily temperature has the highest scores, and the model with maximum temperature has lower scores than the other two. The differences between the scores for the model fit with the mean and maximum temperatures don't differ greatly. This indicates that the predictive performance is best when the maximum temperature is used as the basis for the quadratic temperature term. 

At $5 ^\circ$ c,a $1 ^\circ$ c increase is associated with approximately 116 additional cyclists.

At $15 ^\circ$ c,a $1 ^\circ$ c increase is associated with approximately 161 additional cyclists.

### Seasonal Prediction Accuracy

The month with the lowest RMSE and DS is July, followed closely by June. This indicates that the average value between the predicted and actual values are lowest for June and July, and that M2 best predicts the average value and uncertainty for these months. Cycle numbers are generally higher during these months, and the weather is generally consistent. Furthermore, there are no major holidays falling within these months that could lead to spikes or dips in numbers. 

The month with the highest RMSE and DS is April, followed by March. This indicates poor fit of M2 during March and April, and a worse prediction of the mean cycle count and variation in count. The weather is very volatile during spring, particularly concerning rainfall. For example, a day with higher temperatures but heavy rainfall is likely to see a drop in cyclists, but the data doesn't account for this. This is therefore a limitation in the data. 

The scores for August are considerably higher than the scores for the other Summer months. We believe this is due to the cycling event in August 2022 introduced earlier, which saw massively higher cyclist counts than in other years. More generally, the accuracy may be lower in certain months as the model does not account for holidays and special events, such as the cycling event. This is a limitation of both the model and data, as there could be more data on outside factors, and the model could be updated to account for these cases.

### Long-term trend and extrapolation

We created a table showing the estimated daily and annual change of cyclists over time for each model. 

```{r Trend coefficients from M1-M3}
kable(table3_final, 
             caption = "Trend coefficients from M1-M3") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) 
```

The estimated annual change in daily cyclist count for our final model, M2, is -692. This means that for each year that passes, there is predicted to be 692 fewer daily cyclists. The predicted number of cyclists on the first day of observations (1st January 2020) is 12463. Therefore, the daily cyclist count is estimated to drop by $5.55 \%$ per year. If this trend were to continue, then the daily cyclist count would half within 13 years, and would drop to 0 within 200 years. Additionally, the trend coefficient for M0 has a p-value very close to zero (under 0.01), suggesting statistical significance.  

The estimated trend does change across M1, M2, and M3. The coefficient becomes less negative, but not by much. This indicates that the number of days since the beginning of observations does have an impact on the annual trend, so it is a robust estimator. ***NEED TO CHANGE BECAUSE M3 TREND STUFF IS WRONG*****

The linear trend is a reasonable assumption, since the effect of Covid on cycling is only pronounced during the lockdown, which spanned only a year. As we can see in Figure 9, the shape still follows that of the other years, with a decrease of daily cycling count in winter and highs in summer. Furthermore, some people may have been working from home and therefore not commuting, whilst others may have taken up cycling to avoid public transport. In the absence of other data, it is difficult to predict the effect of Covid. Including a Covid indicator in M3 does(n't) change the trend coefficient by ... ***ADD in***

We created a plot projecting demand forwards to 2035, holding weather and seasonality constant, as follows.

***PLOT***

At the current rate, the mean daily counts under summer weekday conditions would be expected to reach 5000 cyclists a day in ***add***

This extrapolation may not hold in practice because it doesn't take into account rising population counts, a removal of the under 22 free bus scheme, and implementation of further Low Emission Zones, and rising petrol prices.  


### Identifying a poorly fit period

### Original diagnostic plot

# Discussion and limitations

<!--
3-5 paragraphs covering:
  - Whether model assumptions appear satisfied.
  - Key limitations of the model and data.
  - What ideas would improve prediction.
  - Whether findings generalise to other cities or future years.
Do not introduce new results here.
-->

# Appendix - References
$^{1}$ https://mccyclingdiary.blogspot.com/2022/08/london-edinburg-london-lel-2022-edition.html

# Appendix - Code


All analysis code is in `code.R` and is reproduced below.

```{r print-code, echo=TRUE, eval=FALSE, code=readLines("code.R")}
```


