---
title: 'Project 2: Modelling daily cycling demand in Edinburgh'
author: 'Group members: Alice Miller, Saba Sitter, Jake Saunders'
date: 'Word count: [XXX words]'
output:
  html_document:
    df_print: paged
  pdf_document:
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
  word_document: default
header-includes:
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{amsmath}
- \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,    # This hides all code from the main body 
  message   = FALSE, 
  warning   = FALSE
)

source("code.R") # Runs all models and creates tables/plots in the background
```

# Introduction

The "cycle daily" dataset  consists of approximately 2,180 daily observations ranging from 2020 to 2025, sourced from Edinburgh's network of automatic cycling counters, along with local weather and temporal data. This data comes from sensors at specific locations across the city, so it's possible not every cyclist is accounted for. Similarly, some cyclists may pass through multiple sensors and be observed more than once. However, they are a good indicator of how cycling demand varies across different days and weather patterns. For future funding allocation and planning, it is important to know how many cyclists are on the road in order to evaluate whether the infrastructure (cycle lanes, crossings etc) and investment in schemes (eg rentable Voi bikes) need to adapt to meet demand. The modelling goal is to develop a sequence of increasingly complex linear regression models to isolate how variables, such as the day of the week and the temperature, influence cycling demand. The goal is to extrapolate cycling demand so that Edinburgh City Council can optimise infrastructure investment over the next decade to align with their active travel priorities.

# Exploratory data analysis

We began by converting month to an ordered factor and redefining the day of week predictor with explicit levels. We created a new variable (trend) which is the number of days since the first observation (01/01/2020), along wtih binary indicators 'is_covid' and 'is_holiday' which represent the UK Covid lockdown (23/03/2020 - 31/03/2021) and the yearly festive period (24th - 31st December) respectively. 

```{r summary stats exploratory table, echo=FALSE}

kable(summary_stats, 
             digits = 2, 
             caption = "Summary Statistics for count, temp mean, temp min, temp max.")%>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE)
```

We can see from table 1 that the number of daily cyclists varies from $1,264$ to $43,864$, a variation of $\approx 3370 \%$. Intuitively, one would think that a warmer average weather would lead to more cyclists. 

```{r time series, fig.cap="Time series of daily cyclist counts 2020-2025.",fig.height = 3, fig.width = 5}
plot_timeseries
```


Figure 1 shows that there are yearly fluctuations, with peaks and troughs occurring seasonally. Overall, the cyclist count is decreasing over time. There is a visible peak surrounding August 10th 2022, which we investigated. We found there was a London-Edinburgh-London bicycle event taking place over this period with over 2000 riders$^{1}$. Supposing each cyclist passed through multiple sensors over this period, this likely explains the inflated numbers. 

```{r Boxplots of cyclist count by month and day of week, fig.cap="Boxplots of cyclist count by month and day of week.", fig.height = 5, fig.width = 7}
grid.arrange(plot_month,plot_dow, nrow = 2)
```


We can see from Figure 2 that the colder months (October - February) have slightly lower cyclist counts, with the other months remaining relatively constant. There are a lot of outliers, especially in Spring. Observations with unusually high cyclist counts could be attributed to special events such as races, bank holidays, and exceptionally sunny days. Outside of winter, the month doesn't appear to have an impact on the median count. August has a high count outliers, which may come from the combined effect of the Fringe Festival and the cycling event mentioned previously.

The second boxplot in Figure 2 indicates that there tends to be less cyclists on Mondays and Sundays, with the other days appearing relatively constant. 

```{r scatter plot of count vs temp_mean, fig.cap="Scatter plot of Daily cyclist count vs mean temperature.", fig.height = 3, fig.width = 5}
plot_temp_scatter
```

Figure 3 shows a clear positive correlation between temperature mean and daily cyclist count. There are significantly less observations for days with the lowest and highest mean temperatures, making it difficult to extrapolate for these days. There is a large variation in cyclist counts for days with the same temperature, so it is not feasible to predict counts based off of mean temperature alone.  

We noticed that the days with the lowest counts fell between Christmas Eve and New Years Eve, presumably due to less commuters on the road.

```{r plot december, fig.cap="Line Graph of average daily cyclists across the 5 years in decemeber.", fig.height = 3, fig.width = 5}
plot_december

```

We can see from Figure 4 that there is a substantial decrease in cyclist numbers between the specified days. This led us to add a variable "is_holiday" for this period.

The data spans the Covid-19 pandemic, so we added a binary predictor 'is_covid' into the dataframe to indicate whether a day falls within the UK mandated lockdown.

# Baseline model 

We fit an initial linear model, M0, that uses the numeric variables mean temperature, weekend, and month to predict the daily cyclist count. 

The equation for M0 is:

$$\text{count}_i = \beta_0 + \beta_1 \text{temp\_mean}_i + \beta_2 \text{weekend}_i + \beta_3 \text{month}_i + \epsilon_i$$
$$\epsilon_i \sim N(0, \sigma^2)$$

The summary statistics for M0

```{r summary for m0} 

coef_matrix <- summary(m0)$coefficients
coef_matrix <- coef_matrix[, -4]
kable(coef_matrix, 
      digits = 2, 
      caption = "Summary coefficients for model M0") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) 
                
```


Table 2 shows for every $1^\circ$c increase in the mean daily temperature, M0 predicts an increase of 247 cyclists per day, and predicts a 1247 decrease in cyclist count on weekends compared to weekdays. The model suggests that there is a decrease of 247 cyclists for each month passing as M0 uses month as a numeric predictor, therefore ignoring the fact that cyclist counts generally increase in summer and fall in winter.


```{r r vs f and qq plots m0, fig.cap="Diagnostic plots for M0.", fig.height = 4, fig.width = 6}
grid.arrange(p_m0_1,p_m0_2, p_m0_3,p_m0_4, nrow = 2, ncol = 2)
```

The linear assumption seems to be validated as we can see from the Scale-Location plot in Figure 5, and there don't appear to be any points with unreasonable leverage. The Residuals vs Fitted plot shows a mean variance of 0, independent of the fitted values. The residuals show a non-linear pattern, with an arguable 'n' shape above the mean line, indicating there is an underlying non-linear structure yet to be included in our model. The Q-Q plot demonstrates right skewing and heavy tail in the positive quantiles, indicating that M0 is particularly bad at predicting high count observations. Therefore, the normality assumption for linear regression is violated. 

# Proposed Models
## Model 1: Linear temperature with trend and factor variables

$$\text{count}_i = \beta_0 + \beta_1 \text{temp\_mean}_i + \beta_2 \text{trend}_i + \sum_{j=2}^{12} \gamma_j \mathbb{I}(\text{month}_i = j) + \sum_{k=2}^{7} \delta_k \mathbb{I}(\text{dow}_i = k) + \epsilon_i$$

$$\epsilon_i \sim N(0, \sigma^2)$$

The prediction for a Monday in January is represented by $\beta_0$. Every other month or day is measured as the expected difference from the baseline. Unlike M0, M1 takes into account the impact of different days. Interpreting 'month' as a factor variable lets us build a seasonal profile, rather than assuming the month index (1-12) is linear with the count. We drop the numerical 'month' and 'is_weekend' to avoid collinearity with the factor 'month' and 'dow' respectively.


```{r M1 diagnostic plots, fig.cap="Diagnostic Plots for M1", fig.width = 6, fig.height= 4}
grid.arrange(p_m1_1,p_m1_2,p_m1_3,p_m1_4, nrow = 2, ncol = 2)
```

Figure 6 shows that, while M1's 'n' pattern is less pronounced, the sigmoid shape of the mean residual line suggests the linear assumption is still violated. The Q-Q plot shows improved alignment at the lower tail, though a higher leverage point appears. The Scale-Location plot indicates the linear assumption is slightly better justified than in M0. 

## Model M2: Quadratic temperature 

$$\text{count}_i = \beta_0 + \beta_1 \text{temp\_mean}_i + \beta_2 \text{temp\_mean}^2_i + \beta_3 \text{trend}_i + \sum_{j=2}^{12} \gamma_j \mathbb{I}(\text{month}_i = j) + \sum_{k=2}^{7} \delta_k \mathbb{I}(\text{dow}_i = k) + \epsilon_i$$
$$\epsilon_i \sim N(0, \sigma^2)$$

Figure 3 identified a non-linear relationship between temperature and cyclist counts. M2 addresses this by replacing the linear temperature term with a quadratic alternative.

```{r Diagnostic plots for M2, fig.cap="Diagnostic Plots for M2", fig.height = 4, fig.width = 6}
grid.arrange(p_m2_1, p_m2_2, p_m2_3, p_m2_4, nrow = 2, ncol = 2)
```     

Diagnostics in Figure 7 show that the linear assumption remainds justified, though outlier leverage has increased. Since diagnostic plots show minimal change from M1, M2's explanatory power appears similar.

## Model M3: Logarithmic count with nonlinear temperature

$$\log(\text{count}_i + 1) = \beta_0 + \beta_1 \text{temp\_mean}_i + \beta_2 \text{temp\_mean}_i^2 + \beta_3 \text{trend}_i + \sum_{j=2}^{12} \gamma_j \mathbb{I}(\text{month}_i = j) + \sum_{k=2}^{7} \delta_k \mathbb{I}(\text{dow}_i = k) +
$$$$
\beta_4 \text{is\_covid}_i + \beta_5 \text{is\_holiday}_i + \epsilon_i$$
$$\epsilon_i \sim N(0, \sigma^2)$$

M3 applies a log-transform to the count response to mitigate the heavy right-tailed Q-Q plots observed in previous models. Binary predictors 'is_holiday' and 'is_covid' are justified by Figures 8 and 9.


```{r}
plot_holiday + plot_covid
```

The festive period (December 24th-31st) shows significantly lower, more consistent counts. 
The lockdown period shows a more pronounced decline than the overall five-year trend.


```{r Diagnostic plots M3, fig.cap="Diagnostic Plots for M3", fig.height = 4, fig.width = 6}
grid.arrange(p_m3_1, p_m3_2, p_m3_3, p_m3_4, nrow = 2, ncol = 2)
```

Diagnostics for M3 suggest the normality assumption is again justified. However, the Scale-Location plot indicates heteroscedasticity, violating the constant variance assumption. A high-influence point remains visible in the Residuals vs Leverage plot. 

Our final model is M2. 

```{r summary for m2} 
kable(m2_table, 
      digits = 4, 
      caption = "Summary for Model M2") 
                
```

Table 3 shows a baseline of 6,990 cyclists (January Mondays at 0$^{\circ}C$. The positive linear and negative quadratic temperature coefficients confirm that counts rise with temperature but taper off at extremes. January and December see the lowest counts, and all day-of-weeks see higher counts than the baseline. Most predictors are significant ($p < 0.03$), except for October and Sunday.  


# Evaluation
## Statistical Performance

We found the following error scores for each model: root mean squared, mean absolute error, Dawid-Sebastiani score, and interval score. For each year, we found the scores when that year was removed from the training data and used as the test data, and then found the average score for each model. 


```{r Table 1 LOYO CV scores for M0-M3}

kable(table1_final, 
             digits = 2, 
             caption = "Leave-one-year-out CV scores")  %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) 
```

We can see from Table 4 that M2 has the lowest score for RMSE, indicating M2 fits the dataset best. M2 also has the lowest score for MAE, suggesting it is less sensitive to extreme outliers. The low IS demonstrates that M2 produces narrow, specific intervals that include the true count most of the time. M0 has a lower DS than M2 by 0.02, suggesting it is slightly better at predicting the mean and variance of the daily cyclist count.However all of the DS scores are very close, and we established that M0 is the worst model as it doesn't take any quadratic effects of temperature, or the fact that there are factor variables.

This was the basis for us choosing M2 as our final model. 

We then created a table showing the RMSE and DS for each month using M2.

```{r Table 2 RMSE and DS by month}

kable(table2_final, 
             digits = 2, 
             caption = "Table of the RMSE and DS by months")  %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) 
```

## Practical Performance

### Temperature effects and nonlinearlity

The following table fits M2 three times, using temp mean, temp min and temp max in turn as the basis for the temperature terms, and reports the scores. 

```{r M2 fit scores with temp_mean, temp_min, and temp_max}

kable(m2_comparison, 
             digits = 2,
             caption = "Model M2 fitted with daily temp mean, temp min, temp max.") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) 
```

The model with minimum daily temperature has the lowest scores whereas the temp_max model has consistently the highest scores. Generally, the scores from the mean and minimum models are very similar, in particular the DS score is identical up to 2 d.p. We conclude predictive performance is best when temp_min is used as the basis for the temperature terms. This is physically plausible as temperatures are lowest at either side of the working day, when we expect most people to cycle.

At $5^\circ$c, a $1^\circ$c increase is associated with approximately 46 additional cyclists.

At $15^\circ$c, a $1 ^\circ$c increase is associated with approximately 200 fewer cyclists. Which is expected based on the negative quadratic curve at $15^\circ$c in figure 3, thus at higher temperatures, with all other variables constant, we anticipate a decrease in count.


### Seasonal Prediction Accuracy

The lowest RMSE and DS scored month is July, followed closely by June. Thus, the mean differences between the predicted and actual values are lowest for June and July; M2 best predicts the count (including uncertainty) for these months. Cycle numbers are generally higher during these months, and the weather is generally consistent. Furthermore, there are no major holidays falling within these months that could lead to spikes or dips in numbers. 

The highest RMSE and DS scores come from April, followed by March. This indicates a poor fit of M2, and a worse prediction of the mean cycle count and variation in count.
Some of the March volatility may come from the March 2020 stratum, when lockdown was dramatically enforced and restrictions were most severe and respected. The weather, especially rainfall is most volatile during spring. For instance, a day with higher temperatures but heavy rainfall is likely to result in less cyclists. We hypothesise rainfall may be a greater influence than temperature but but the data doesn't account for this.

The August scores are considerably higher than other summer months. We hypotheise this is due to the cycling event in August 2022 introduced earlier, which saw exceptionally high cyclist count compared to other years. The prediction accuracy is limited since the data does not account for holidays and special events, such as the cycling event or Fringe Festival. This is a limitation of both the model and data, since with the improved data, a more comprehensive model could be built.

### Long-term trend and extrapolation

$\textbf{Annual Growth Rate}$ 

We created a table showing the estimated daily and annual change of cyclists over time for each model. 

```{r Trend coefficients from M1-M3}
kable(table3_final, 
             caption = "Trend coefficients from M1-M3") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) 
```

The estimated annual change in daily cyclist count for our final model M2 is -683. Thus for each year passed, we predict 683 fewer daily cyclists. The predicted number of cyclists on the first day of observations (1st January 2020) is 12,463. Therefore, the daily cyclist count is estimated to drop by $5.48 \%$ per year. If this trend were to continue, then the daily cyclist count would half within 9 to 10 years.

Additionally, the trend coefficient for all of the models have extremely small p-values (less than 0.01). This provides strong evidence that the downward trend observed in the daily cyclist count is not random, but is instead an observable decline in cycling demand over the 5 year period. 

$\textbf{Model consistency}$ 

M1 and M2 produced similar results, with annual declines of -692 and -683 respectively. This suggests that the trend estimate is robust even when accounting for the quadratic effect of temperature, the month and days of week factor variables. M3 shows a steeper decline of -887 cyclists per year, approximately 30% lower than M2.

This difference is likely due to the logarithmic scale in M3, which is more sensitive to relative percentage changes. M3 includes indicators for the Covid and festive holidays, so accounts for the variance associated with each to isolate the downward trend. 

Despite variations in magnitude, we observe the negative trend coefficient that is highly significant for all models demonstrates and thus the variable is a robust addition to model

$\textbf{Covid Distortion}$ 

The linear trend is a reasonable baseline assumption for the 2020-2025 period, as the most disruptive impact of the lockdown was concentrated within a one-year window. As shown in Figure 8, the fundamental yearly structure with winter troughs and summer peaks remained consistent. Additionally, those working from home would not be commuting, whilst others may have taken up cycling to avoid public transport. Despite this, Figure 8 suggests that these fluctuations did not fundamentally affect the long-term trajectory of the data. 

Including a Covid indicator in M3 does change the trend coefficient significantly. In our final model, M2, which does not account specifically for Covid, the daily decline was -1.872 cyclists. Once the Covid indicator was added, the daily trend coefficient shifted to -2.431. This suggests the underlying daily decline is steeper than the other models suggest. To ensure that the shift wasn't purely due to the 'is_holiday' predictor, we fitted M3 without it. The findings remained consistent, confirming that the lockdown period is the primary factor distorting the long-term linear trend estimate. 


$\textbf{Extrapolation}$ 

We projected demand to 2035 using Model M2, holding weather and seasonality constant for a Wednesday in July at the mean temperature. July was chosen as it represents peak annual demand, providing a baseline for infrastructure requirements.

```{r plot extrapolation, fig.cap="Plot projecting demand forward to 2035, for a Wednesday in July of mean temperature", fig.height = 3, fig.width = 5}
extrapolation_plot
```

While July typically sees peak cycling volumes, the model indicates a downward trajectory. As shown in Figure 11, the projected daily count starts at approximately 15,000 in 2020 and falls to roughly 5000 by 2035. At this rate, summer weekday counts are expected to drop below 10,000 on the 2nd of February 2027, and reach 5000 by the 26th of May 2034. 

This extrapolation may not hold if policy or economic factors shift. Expanding Low Emission Zones or rising petrol prices may disincentivise driving, while removing the under-22 Free Bus Scheme might encourage younger demographics to return to cycling. Additionally, Edinburgh's growing population could sustain a minimum demand level that a linear trend does not take into account.


### Identifying a poorly fit period

```{r plot residuals vs time, fig.cap="Plot of Residuals vs Time, coloured by year", fig.height = 3, fig.width = 6}
plot_residuals_years
```

Figure 12 shows that Model M2 performs poorly during April 2022, where the actual cyclist count exceeded predictions by over 10,000. During this period, the model recorded a mean residual of +8,033 and an RMSE of 10,528 - over double the model's LOYO RMSE. This systematic under-prediction likely stems from the Edinburgh International Science Festival (9th-24th April 2022), which attracted 50,000 participants and emphasised climate urgency$^{2}$. Following two years of lockdown cancellations, this event created a demand spike that the baseline model could not anticipate.

### Original diagnostic plot

We designed a Cook's Distance plot over time to show the degree of influence that each month in each year exerts.

```{r cooks distance plot over time, fig.cap="Plot of Cook's distance over time", fig.height = 4, fig.width = 6}
faceted_cooks_plot
```

Figure 13 identifies three points in August 2022 as having the highest influence on the model. This is in line with Figure 1, likely due to Edinburgh-London-Edinburgh race. Additional high-influence points in Spring 2022 coincide with the final lifting of Covid restrictions, a period where both commuting and leisure cycling surged. From a planning perspective, these findings highlight a critical limitation: while M2 is effective for baseline trends, it can't account for exceptional events causing spikes. Planners must prepare for this, particularly during the Fringe Festival and major bike races, by adjusting infrastructure and safety measures beyond what the standard model suggests.


# Discussion and limitations

The model assumptions generally appear satisfied, as shown in Figure 7. It is worth noting the point of high leverage, shown in Figure 13 to correspond to the August 2022 cycle race. While this outlier influences the model, the quadratic temperature terms and daily trend effectively capture the data's primary structure.

A key limitation is the five-year observation window, which was heavily impacted by COVID-19. Access to pre-pandemic years would better encapsulate seasonality and highlight effects the model may have missed, such as the 2020 Fringe Festival closure. Furthermore, the lack of data on rainfall and wind speed is a constraint; interaction effects between these variables and temperature would likely predict cyclist counts more accurately than mean temperature alone.

Predictions could be improved by adding granular variables for the Fringe Festival and school holidays. Additionally, the transition out of lockdown was not a discrete event; including specific data on when various restrictions were eased would be good. We also lack information regarding the "strategic locations" of the counters. Adjusting for local data and population density would help account for demographic shifts influencing volume.

General commuting patterns may be generalised to similar cities, but Edinburgh-specific factors like the Fringe are unique. It is difficult to extrapolate long-term trends from only five years of data. Although the downward trend is statistically clear, the lack of more diverse predictors makes it difficult to say with certainty if this decline will persist or if demand will stabilise.



# Appendix - References
$^{1}$ https://mccyclingdiary.blogspot.com/2022/08/london-edinburg-london-lel-2022-edition.html
$^{2}$ https://theedinburghreporter.co.uk/2022/04/its-a-wrap-as-esf-2022-closes-after-another-successful-edition/


# Appendix - Code


All analysis code is in `code.R` and is reproduced below.

```{r print-code, echo=TRUE, eval=FALSE, code=readLines("code.R")}
```


