---
title: 'Project 2: Modelling daily cycling demand in Edinburgh'
author: 'Group members: Alice Miller, Saba Sitter, Jake Saunders'
date: 'Word count: [XXX words]'
output:
  pdf_document:
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
  html_document:
    df_print: paged
header-includes:
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{amsmath}
- \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,    # This hides all code from the main body 
  message   = FALSE, 
  warning   = FALSE
)
source("code.R") # Runs all models and creates tables/plots in the background
```

# Introduction

The "cycle daily" dataset  consists of approximately 2,180 daily observations ranging from 2020 to 2025, sourced from Edinburgh's network of automatic cycling counters, along with local weather and temporal data. This data comes from sensors at specific locations across the city, so it's possible not every cyclist is accounted for. However, they are a good indicator of how cycling demand varies across different days and weather patterns. Edinburgh relies heavily on public transport for its residents and tourists to navigate the city. For future funding allocation and planning, it is important to know how many cyclists are on the road in order to evaluate whether the infrastructure (cycle lanes, crossings etc) and investment in schemes (rentable Voi bikes) need to adapt to meet demand. The modelling goal is to develop a sequence of increasingly complex linear regression models to isolate how temporal and weather variables, such as the day of the week and the temperature, influence cycling demand. Ultimately, the goal is to extrapolate cycling demand to the future so that Edinburgh City Council and Transport Scotland can optimise infrastructure investment over the next decade to align with their active travel priorities.


# Exploratory data analysis

To begin investigating the impact of weather on the number of daily cyclists, we begun with some exploratory data analysis. Table 1 displays the summary statistics for the weather data and number of daily cyclists is as follows:

## Summary Statistics table for count, temp_mean, temp_min, temp_max

```{r summary stats exploratory table, echo=FALSE}

knitr::kable(summary_stats, 
             digits = 2, 
             caption = "Summary Statistics for count, temp mean, temp min, temp max")
```
We can see that the number of daily cyclists varies from $1,264$ to $43,864$, an increase of $\approx 3470 \%$, and the variation is very high. Intuitively, one would think that a warmer average weather would lead to more cyclists. 

We then plotted a time series of the daily cyclist count from 2020 to 2025 (figure 1). 

## Time series of daily counts 2020-2025 with a smoother.

Below is ...

```{r time series, fig.cap="Time series of daily counts 2020-2025 with a smoother"}
plot_timeseries
```


We can see from the time series that there are yearly fluctuations, with peaks and troughs occuring yearly. Overall, the cyclist count seems to be decreasing. There is a visible peak in 2022, which we investigate. We then created boxplots of the daily cyclist count by month and by the day of week. 


## Boxplots of count by month and by day of week.
```{r boxplots of count by month and day of week, fig.cap="Boxplots of count by month and by day of week"}
grid.arrange(plot_month,plot_dow, nrow = 2)
```


We can see from these plots that the winter months have slightly lower cyclist counts, with the other months remaining relatively constant. This indicates that the month does not have a tremendous impact on the number of daily cyclists outside of January and December. From the second boxplot, there appear to be less cyclists on Mondays and Sundays, with the other days again appearing relatively constant. 

## Scatter plot of count vs temp_mean.

```{r scatter plot of count vs temp_mean, fig.cap="Scatter plot of count vs temperature mean"}
plot_temp_scatter
```


# Baseline model M0

<!--
Required content:
  1. State M0 mathematically using a numbered equation.
  2. Coefficient table (use kable, not raw output).
  3. Residuals vs fitted and normal QQ diagnostic plots, side by side.
  4. Brief assessment: what does M0 get right and wrong? 
     What motivates M1?
-->

# Proposed models M1--M3

<!--
For each model:
  - State the model mathematically.
  - Justify every new term BEFORE showing results.
  - Show diagnostic plots and comment on improvement over the previous model.
-->

## Model M1: Trend and factor 
## Model M2: Nonlinear temperature
## Model M3: [your extension title]

# Evaluation
## Statistical Performance

<!--
Required content:
  1. Table 1: LOYO CV scores for M0-M3.
  2. Table 2: CV RMSE and DS by month for your final model.
  3. Discussion: do the scores agree on which model is best? 
     If not, explain why and state your final model choice explicitly.
-->

## Practical Performance

<!--
One subsection per planning question (5.1-5.5 from the assignment brief).
Each must include mandatory fixed-format numbers and real-world interpretation.
-->

# Discussion and limitations

<!--
3-5 paragraphs covering:
  - Whether model assumptions appear satisfied.
  - Key limitations of the model and data.
  - What ideas would improve prediction.
  - Whether findings generalise to other cities or future years.
Do not introduce new results here.
-->

# Appendix: Code

All analysis code is in `code.R` and is reproduced below.

```{r print-code, echo=TRUE, eval=FALSE, code=readLines("code.R")}
```


