---
title: 'Project 2: Modelling daily cycling demand in Edinburgh'
author: 'Group members: Alice Miller, Saba Sitter, Jake Saunders'
date: 'Word count: [XXX words]'
output:
  pdf_document:
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{amsmath}
- \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,    # This hides all code from the main body 
  message   = FALSE, 
  warning   = FALSE
)

source("code.R") # Runs all models and creates tables/plots in the background
```

# Introduction

The "cycle daily" dataset  consists of approximately 2,180 daily observations ranging from 2020 to 2025, sourced from Edinburgh's network of automatic cycling counters, along with local weather and temporal data. This data comes from sensors at specific locations across the city, so it's possible not every cyclist is accounted for. Similarly, some cyclists may pass through multiple sensors in a day and be observed more than once. However, they are a good indicator of how cycling demand varies across different days and weather patterns. Edinburgh relies heavily on public transport for its residents and tourists to navigate the city. For future funding allocation and planning, it is important to know how many cyclists are on the road in order to evaluate whether the infrastructure (cycle lanes, crossings etc) and investment in schemes (eg rentable Voi bikes) need to adapt to meet demand. The modelling goal is to develop a sequence of increasingly complex linear regression models to isolate how temporal and weather variables, such as the day of the week and the temperature, influence cycling demand. Ultimately, the goal is to extrapolate cycling demand so that Edinburgh City Council and Transport Scotland can optimise infrastructure investment over the next decade to align with their active travel priorities.


# Exploratory data analysis

To begin investigating the impact of weather on the number of daily cyclists, we begun with some exploratory data analysis. This involved converting month to an ordered factor, redefining the day of week predictor with explicit levels. Additionally, we created a new variable 'trend' which symbolises the number of days since the first observation (01/01/2020), along wtih binary indicators 'is_covid' and 'is_holiday' which represent the UK Covid lockdown (23/03/2020 - 31/03/2021) and the yearly festive period (24th - 31st December) respectively. 

We initially found the summary statistics for the number of daily cyclists and the weather data.

```{r summary stats exploratory table, echo=FALSE}

kable(summary_stats, 
             digits = 2, 
             caption = "Summary Statistics for count, temp mean, temp min, temp max.")%>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE)
```

We can see from table 1 that the number of daily cyclists varies from $1,264$ to $43,864$, a variation of $\approx 3370 \%$. Intuitively, one would think that a warmer average weather would lead to more cyclists. 

We then plotted a time series of the daily cyclist count over the period of observations. 

```{r time series, fig.cap="Time series of daily cyclist counts 2020-2025.",fig.height = 2, fig.width = 4}
plot_timeseries
```


We can see from Figure 1 that there are yearly fluctuations, with peaks and troughs occurring seasonally. Overall, the cyclist count seems to be decreasing over time. There is a visible peak on and surrounding the 10th of August 2022, which we investigated. We found that there was a London-Edinburgh-London bicycle event taking place over this period with over 2000 riders$^{1}$. Supposing each cyclist passed through multiple sensors over the course of the days, this likely explains the inflated numbers. 

We then created boxplots of the daily cyclist count by month and by the day of week to observe whether they had any significant impact on the number of daily cyclists. 


```{r Boxplots of cyclist count by month and day of week, fig.cap="Boxplots of cyclist count by month and day of week.", fig.height = 3, fig.width = 4}
grid.arrange(plot_month,plot_dow, nrow = 2)
```


We can see from Figure 2 that the colder months (October - February) have slightly lower cyclist counts, with the other months remaining relatively constant. There are a lot of outliers, especially in Spring. Observations with unusually high cyclist counts could be attributed to special events such as races, bank holidays, and exceptionally sunny days. Outside of winter, the month doesn't appear to have a tremendous impact on the median daily cyclist count. The second boxplot in Figure 2 indicates that there tends to be less cyclists on Mondays and Sundays, with the other days appearing relatively constant. 

We then plotted the daily cyclist count against the mean temperature, with the red and blue lines indicating regression and quadratic relationship respectively as seen in Figure 3. 

```{r scatter plot of count vs temp_mean, fig.cap="Scatter plot of Daily cyclist count vs mean temperature.", fig.height = 2, fig.width = 4}
plot_temp_scatter
```

There is a clear positive correlation between temperature mean and daily cyclist count, which is to be expected. There are significantly less observations for the days with the lowest and highest mean temperatures, so there is not a clear trend that we can see for those days. There is a large variation in cyclist counts for days with the same temperature, so it is not feasible to predict counts based off of mean temperature alone.  

We noticed that the days with the lowest counts fell between Christmas Eve and New Years Eve, presumably due to less commuters on the road. We created a graph to illustrate this over the course of the 5 years. 

```{r plot december, fig.cap="Line Graph of average daily cyclists across the 5 years in decemeber.", fig.height = 2, fig.width = 4}
plot_december

```

We can see from Figure 4 that there is a substantial decrease in cyclist numbers between the specified days. This led us to add whether a day falls between the 24th and 31st of December in our final model as 'is_holiday'.

The data spans the Covid-19 pandemic, so we added a binary predictor 'is_covid' into the dataframe to indicate whether a day falls within the UK mandated lockdown, which we later investigated.

# Baseline model 

We fit an initial linear model, M0, that uses the numeric variables mean temperature, weekend, and month to predict the daily cyclist count. 

The equation for M0 is:

$$\text{count}_i = \beta_0 + \beta_1 \text{temp\_mean}_i + \beta_2 \text{weekend}_i + \beta_3 \text{month}_i + \epsilon_i$$
$$\epsilon_i \sim N(0, \sigma^2)$$

The summary statistics for M0

```{r summary for m0} 

coef_matrix <- summary(m0)$coefficients
coef_matrix <- coef_matrix[, -4]
kable(coef_matrix, 
      digits = 2, 
      caption = "Summary coefficients for model M0") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) 
                
```

For every 1 degree increase in the mean daily temperature, M0 predicts an increase of 247 cyclists per day, and predicts a 1247 decrease in cyclist count on weekends compared to weekdays. Currently, the model suggests that there is a decrease of 247 cyclists for each month passing. This is because M0 uses month as a numeric predictor and therefore it ignores the fact that cyclist counts generally increase in summer and fall in winter.

We then assessed the diagnostic plots for M0.

```{r r vs f and qq plots m0, fig.cap="Diagnostic plots for M0.", fig.height = 2, fig.width = 8}
grid.arrange(p_m0_1,p_m0_2, p_m0_3,p_m0_4, ncol = 4)
```

The linear assumption seems to be validated as we can see from the Scale-Location plot in Figure 6, and there don't appear to be any points with unreasonable leverage. The Residuals vs Fitted plot in Figure 5 shows a mean variance of 0, independent of the fitted values. The residuals show a non-linear pattern, with an arguable 'n' shape above the mean line, indicating there is an underlying non-linear structure yet to be included in our model. The Q-Q plot in Figure 5 demonstrates right skewing and heavy tail in the positive quantiles, indicating that M0 is particularly bad at predicting high count observations. Therefore, the normality assumption for linear regression is violated. 

# Proposed Models
## Model 1: Linear temperature with trend and factor variables

We created a further model, M1, which adds in the trend variable as a predictor to capture the drift in cycling levels over 2020-2025, and interprets day of week and month as factor variables rather than numeric variables.

Our proposed equation for M1 (identifiable form) is:
$$\text{count}_i = \beta_0 + \beta_1 \text{temp\_mean}_i + \beta_2 \text{trend}_i + \sum_{j=2}^{12} \gamma_j \mathbb{I}(\text{month}_i = j) + \sum_{k=2}^{7} \delta_k \mathbb{I}(\text{dow}_i = k) + \epsilon_i$$

$$\epsilon_i \sim N(0, \sigma^2)$$

The prediction for day 1 (Monday) and month 1 (Jan) are represented by the intercept term $\beta_0$
Every other month or day is measured as the expected difference in count from the baseline. M1 is an improvement over M0 because M1 takes into account the impact of different weekdays rather than simplistic weekend or non weekend effects. Changing 'month' to a factor variable allows for building a seasonal profile rather than assuming the month index (1-12) is linear with the count.
We omit the weekend term and numerical version of 'month' to avoid collinearity with their 'factor' versions and because the covariate 'dow' is collinear with 'weekend'.


```{r side by side comparison of M0 and M1, fig.cap="Diagnostic Plots for M1", fig.width = 8, fig.height = 2}
grid.arrange(p_m1_1,p_m1_2,p_m1_3,p_m1_4, nrow = 1, ncol = 4)
```

We can see from the Residuals vs Fitted Values plot in Figure 6 that, despite the 'n' pattern about the mean being slightly less pronounced, the mean residual line follows a slight sigmoid shape, suggesting the linear assumption is still violated for M1. The Q-Q plot looks slightly more aligned to the normal line at the lower tail in comparison to M1, suggesting that the prediction for the days with lower cyclist counts has improved. There is now however a point with very large leverage as seen in the Residuals vs Leverage plot. The linear assumption appears to be slightly more justified than with M0, as we can see from the Scale-Location plot.

## Model M2: Quadratic temperature 

Our proposed equation for M2 (identifiable form) is:
$$\text{count}_i = \beta_0 + \beta_1 \text{temp\_mean}_i + \beta_2 \text{temp\_mean}^2_i + \beta_3 \text{trend}_i + \sum_{j=2}^{12} \gamma_j \mathbb{I}(\text{month}_i = j) + \sum_{k=2}^{7} \delta_k \mathbb{I}(\text{dow}_i = k) + \epsilon_i$$
$$\epsilon_i \sim N(0, \sigma^2)$$

We can infer from Figure 3 that there is an underlying non-linear relationship between the mean temperature and the daily cyclist count. Hence, replacing the linear 'mean temperature' term with a nonlinear alternative term, the underlying structure may be better explained by the model. We can evaluate this using the following diagnostic plots for M2.

```{r 1st_Diagnostics_of_M1_M2, fig.cap="Diagnostic Plots for M2", fig.height = 2, fig.width = 8}
grid.arrange(p_m2_1, p_m2_2, p_m2_3, p_m2_4, nrow = 1, ncol = 4)
```     

The linear assumption still appears to be justified and there is still a point with massive influence. There is not much difference from M1 in the diagnostic plots, suggesting that the explanatory power of M2 is not improved upon M1.


## Model M3: Logarithmic count with nonlinear temperature

The expression is: 
$$\log(\text{count}_i + 1) = \beta_0 + \beta_1 \text{temp\_mean}_i + \beta_2 \text{temp\_mean}_i^2 + \beta_3 \text{trend}_i + \sum_{j=2}^{12} \gamma_j \mathbb{I}(\text{month}_i = j) + \sum_{k=2}^{7} \delta_k \mathbb{I}(\text{dow}_i = k) +
$$$$
\beta_4 \text{is\_covid}_i + \beta_5 \text{is\_holiday}_i + \epsilon_i$$
$$\epsilon_i \sim N(0, \sigma^2)$$

To mediate the heavy right tailed Q-Q plots seen in the previous models, we apply the log 
transform to the count response. We added in binary predictor variables - 'is_holiday' and 'is_covid' - introduced and defined during exploratory data analysis. 


```{r Boxplot of holiday, fig.cap="Boxplot of Holiday variable (Dec 24th - 31st)", fig.height = 2, fig.width = 4}
plot_holiday

# holiday boxplots
plot_holiday <- ggplot(cycle_daily_df, aes(x = factor(is_holiday), y = count)) +
  geom_boxplot() +
  labs(title = "Impact of Festive Period on Daily Cycle Count",
       x = "Is Holiday (0=No, 1=Yes)", y = "Cycle Count") +
  theme_minimal()

```

```{r Impact of adding covid, fig.cap="Trend Lines of cyclist counts with and without Covid variable.", fig.width = 4, fig.height = 2}
plot_covid
# COVID variable
plot_covid <- ggplot(cycle_daily_df, aes(x = date, y = count, color = factor(is_covid))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(title = "Cycling Trends: Non Covid vs. Covid Era") +
  theme_minimal()

```


```{r M3 compared to M2}
grid.arrange(p_m2_1, p_m2_2, p_m3_1, p_m3_2, nrow = 2, ncol = 2)
grid.arrange(p_m2_3, p_m2_4, p_m3_3, p_m3_4, nrow = 2, ncol = 2)
```



# Evaluation
## Statistical Performance

<!--
Required content:
  1. Table 1: LOYO CV scores for M0-M3.
  2. Table 2: CV RMSE and DS by month for your final model.
  3. Discussion: do the scores agree on which model is best? 
     If not, explain why and state your final model choice explicitly.
-->

Table of the Leave-one-year-out CV scores.

```{r Table 1 LOYO CV scores for M0-M3}

knitr::kable(table1_final, 
             digits = 2, 
             caption = "Table of the Leave-one-year-out CV scores")
```

Table of the RMSE and DS by month

```{r Table 2 of the RMSE and DS by month}

knitr::kable(table2_final, 
             digits = 2, 
             caption = "Table of the RMSE and DS by months")
```

## Practical Performance

<!--
One subsection per planning question (5.1-5.5 from the assignment brief).
Each must include mandatory fixed-format numbers and real-world interpretation.
-->

### 5.1 Temperature effects and nonlinearlity

The table shows the different score values for model 2, but using 3 different temperature variables:temp mean, temp min and temp max.

```{r M2 fit with temp_mean, temp_min, and temp_max}

knitr::kable(m2_comparison, 
             digits = 2,
             caption = "Model M2 fitted with daily temp mean, temp min, temp max.")
```

The model with temp max in the nonlinear term has the best predictive performance because it has the lowest value for each score.

At $5 ^\circ$ c,a $1 ^\circ$ c increase is associated with approximately 108 additional cyclists.

At $15 ^\circ$ c,a $1 ^\circ$ c increase is associated with approximately 166 additional cyclists.


# Discussion and limitations

<!--
3-5 paragraphs covering:
  - Whether model assumptions appear satisfied.
  - Key limitations of the model and data.
  - What ideas would improve prediction.
  - Whether findings generalise to other cities or future years.
Do not introduce new results here.
-->

# Appendix - References
$^{1}$ https://mccyclingdiary.blogspot.com/2022/08/london-edinburg-london-lel-2022-edition.html

# Appendix - Code


All analysis code is in `code.R` and is reproduced below.

```{r print-code, echo=TRUE, eval=FALSE, code=readLines("code.R")}
```


