---
title: 'Project 2: Modelling daily cycling demand in Edinburgh'
author: 'Group members: Alice Miller, Saba Sitter, Jake Saunders'
date: 'Word count: [XXX words]'
output:
  pdf_document:
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
  html_document:
    df_print: paged
header-includes:
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{amsmath}
- \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,    # This hides all code from the main body 
  message   = FALSE, 
  warning   = FALSE
)
source("code.R") # Runs all models and creates tables/plots in the background
```

# Introduction

<!--
Required content (remove before submission):
3-5 sentences covering:
  - What the dataset is and where it comes from
  - What the modelling goal is
  - Why this is useful for transport planning in Edinburgh
Do NOT describe the report structure here.
-->
The "cycle daily" dataset contains the number of cyclists per day across Edinburgh. This data comes from sensors at specific locations therefore it is possible they don't include every cyclist however they do give a good general idea of the trend against different days of the week, seasons, years and temperature. Our aim with the models is: to see if we can accurately predict the number of cyclists over a given period based on different factors, to see which factors have the most influence, to see if there is any other contributing effects to take into account and to extrapolate demand for the next 10 years.
Edinburgh is a very dense city that relies heavily on public transport for its residents and tourists to get around. For future funding allocation and planning it is important to know how many cyclists are on the road, to see if the infrastructure (cycle lanes, crossings etc) need to increase to meet demand. It also lets the council know if they should invest more in other schemes such as the Voi bikes. 

# Exploratory data analysis

<!--
Required content:
  1. Summary statistics table for count, temp_mean, temp_min, temp_max.
  2. Time series of daily counts 2020-2025 with a smoother.
  3. Boxplots of count by month and by day of week.
  4. Scatter plot of count vs temp_mean.
  5. At least one further plot of your choice highlighting a notable feature.
Every figure must have a caption and be explicitly discussed in the text.
-->

```{r q2-model, echo=FALSE}

knitr::kable(summary_stats, 
             digits = 4, 
             caption = "Summary statistics for daily cyclist counts and temperature variables (2020-2025)")
```

# Baseline model M0

<!--
Required content:
  1. State M0 mathematically using a numbered equation.
  2. Coefficient table (use kable, not raw output).
  3. Residuals vs fitted and normal QQ diagnostic plots, side by side.
  4. Brief assessment: what does M0 get right and wrong? 
     What motivates M1?
-->

# Proposed models M1--M3

<!--
For each model:
  - State the model mathematically.
  - Justify every new term BEFORE showing results.
  - Show diagnostic plots and comment on improvement over the previous model.
-->

## Model M1: Trend and factor 
## Model M2: Nonlinear temperature
## Model M3: [your extension title]

# Evaluation
## Statistical Performance

<!--
Required content:
  1. Table 1: LOYO CV scores for M0-M3.
  2. Table 2: CV RMSE and DS by month for your final model.
  3. Discussion: do the scores agree on which model is best? 
     If not, explain why and state your final model choice explicitly.
-->

## Practical Performance

<!--
One subsection per planning question (5.1-5.5 from the assignment brief).
Each must include mandatory fixed-format numbers and real-world interpretation.
-->

# Discussion and limitations

<!--
3-5 paragraphs covering:
  - Whether model assumptions appear satisfied.
  - Key limitations of the model and data.
  - What ideas would improve prediction.
  - Whether findings generalise to other cities or future years.
Do not introduce new results here.
-->

# Appendix: Code

All analysis code is in `code.R` and is reproduced below.

```{r print-code, echo=TRUE, eval=FALSE, code=readLines("code.R")}
```


