---
title: 'Project 2: Modelling daily cycling demand in Edinburgh'
author: 'Group members: Alice Miller, Saba Sitter, Jake Saunders'
date: 'Word count: [XXX words]'
output:
  pdf_document:
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
  html_document:
    df_print: paged
header-includes:
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{amsmath}
- \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,    # This hides all code from the main body 
  message   = FALSE, 
  warning   = FALSE
)
source("code.R") # Runs all models and creates tables/plots in the background
```

# Introduction

The "cycle daily" dataset  consists of approximately 2,180 daily observations ranging from 2020 to 2025, sourced from Edinburgh's network of automatic cycling counters, along with local weather and temporal data. This data comes from sensors at specific locations across the city, so it's possible not every cyclist is accounted for. Similarly, some cyclists may pass through multiple sensors in a day and be observed more than once. However, they are a good indicator of how cycling demand varies across different days and weather patterns. Edinburgh relies heavily on public transport for its residents and tourists to navigate the city. For future funding allocation and planning, it is important to know how many cyclists are on the road in order to evaluate whether the infrastructure (cycle lanes, crossings etc) and investment in schemes (eg rentable Voi bikes) need to adapt to meet demand. The modelling goal is to develop a sequence of increasingly complex linear regression models to isolate how temporal and weather variables, such as the day of the week and the temperature, influence cycling demand. Ultimately, the goal is to extrapolate cycling demand so that Edinburgh City Council and Transport Scotland can optimise infrastructure investment over the next decade to align with their active travel priorities.


# Exploratory data analysis

To begin investigating the impact of weather on the number of daily cyclists, we begun with some exploratory data analysis. This involved converting month to an ordered factor, redefining the day of week predictor with explicit levels. Additionally, we created a new variable 'trend' which symbolises the number of days since the first observation (01/01/2020), along wtih binary indicators 'is_covid' and 'is_holiday' which represent the UK Covid lockdown (23/03/2020 - 31/03/2021) and the yearly festive period (24th - 31st December) respectively. 

We initially found the summary statistics for the number of daily cyclists and the weather data.

```{r summary stats exploratory table, echo=FALSE}

knitr::kable(summary_stats, 
             digits = 2, 
             caption = "Summary Statistics for count, temp mean, temp min, temp max")
```

We can see that the number of daily cyclists varies from $1,264$ to $43,864$, a variation of $\approx 3370 \%$. Intuitively, one would think that a warmer average weather would lead to more cyclists. 

We then plotted a time series of the daily cyclist count over the period of observations. 

```{r time series, fig.cap="Time series of daily cyclist counts 2020-2025"}
plot_timeseries
```


We can see from Figure X that there are yearly fluctuations, with peaks and troughs occurring seasonally. Overall, the cyclist count seems to be decreasing over time. There is a visible peak on and surrounding the 10th of August 2022, which we investigated. We found that there was a London-Edinburgh-London bicycle event taking place over this period with over 2000 riders$^{1}$. Supposing each cyclist passed through multiple sensors over the course of the days, this likely explains the inflated numbers. 

We then created boxplots of the daily cyclist count by month and by the day of week to observe whether they had any significant impact on the number of daily cyclists. 


```{r Boxplots of cyclist count by month and day of week, fig.cap="Boxplots of cyclist count by month and day of week"}
grid.arrange(plot_month,plot_dow, nrow = 2)
```


We can see from Figure 2 that the colder months (October - February) have slightly lower cyclist counts, with the other months remaining relatively constant. There are a lot of outliers, especially in Spring. Observations with unusually high cyclist counts could be attributed to special events such as races, bank holidays, and exceptionally sunny days. Outside of winter, the month doesn't appear to have a tremendous impact on the median daily cyclist count. 

The second boxplot indicates that there tends to be less cyclists on Mondays and Sundays, with the other days appearing relatively constant. 

We then plotted the daily cyclist count against the mean temperature, with red and blue lines indicating regression and quadratic relationship respectively. 

```{r scatter plot of count vs temp_mean, fig.cap="Daily cyclist count vs mean temperature"}
plot_temp_scatter
```

There is a visible positive correlation between temperature mean and daily cyclist count, which was to be expected. There are less observations for the days with the lowest and highest mean temperatures, so there is not a clear trend that we can see for those days. There is a large variation in cyclist counts for days with the same temperature, so it is not feasible to predict counts based off of mean temperature alone.  

We noticed dips in cyclist numbers between Christmas Eve and New Years Eve, presumably due to less commuters on the road. We created the following graph to illustrate this over the course of the 5 years. 

```{r plot showing december, fig.cap="Daily cyclist count vs mean temperature"}
plot_december
```

Clearly, there is a substantial decrease in cyclist numbers between the specified days. This led us to add whether a day falls between the 24th and 31st of December in our final model as 'is_holiday'.

We also created a chart that demonstrates the effect of whether a day is lockdown on daily cyclist counts.

# Baseline model 

We fit an initial model, M0, that uses the mean temperature, along with the month and whether it is a weekend in order to predict the daily cyclist count.

The equation for M0 is:

#$$\text{count}_i = \beta_0 + \beta_1 \text{temp_mean}_i + \beta_2 \text{weekend}_i + \beta_3 \text{month}_i + \epsilon_i$$
#$$\epsilon_i \sim N(0, \sigma^2)$$

The following table shows the summary statistics for M0:

```{r summary for m0, echo=FALSE} 
##notice the p values all 0 because of rounding

knitr::kable(summary(m0)$coefficients, 
             caption = "Summary of M0")
```
For every 1 degree increase in the mean daily temperature, the model predicts an increase of 247 cyclists per day, and predicts a 1247 decrease in cyclist count on weekends compared to weekdays. Since M0 uses month as a numeric predictor, it ignores the fact that cyclist counts generally increase in summer and fall in winter. Currently, the model suggests that there is a decrease of 247 cyclists for each month passing.

We then assessed the diagnostic plots for M0. 

```{r r vs f and qq plots m0, fig.cap="Diagnostic plots for M0"}
grid.arrange(p_m0_1,p_m0_2, ncol = 2)
```
The Normal

Model 0 (m0) shows a mean variance of 0, independent of the fitted values.
The residuals show a non-linear pattern, with an arguable 'n' shape above the mean line, indicating there is an underlying non-linear structure yet to be included in our model.

The QQ plot demonstrates right skewing and heavy tail in the positive quantiles - 
  thus the normality assumption for linear regression is violated.
This indicates the model is particularly bad at predicting high count observations.


# Model M1

We created a further model, M1, which adds in the trend variable as a predictor to capture the drift in cycling levels over 2020-2025.

Our proposed equation for M1 is:
$$\text{count}_i = \beta_0 + \beta_1 \text{temp_mean}_i + \beta_2 \text{trend}_i + \beta_k \text{month}_i + \beta_j \text{dow}_i +  \epsilon_i$$ 
$$\epsilon_i \sim N(0, \sigma^2), k \in \{1,...,12 \}, j \in \{1,...,7 \} $$

Rational for investigating m1 over m0:
m1 has the granularity for the effect of different weekdays rather than mere weekend or non weekend effects.
Changing 'month' to a factor variable allows for building a seasonal profile rather than assuming the month index (1-12) is linear with the count.

We omit the weekend term and numerical version of month to avoid collinearity with their 'factor' versions and because the covariate dow is collinear with weekend.


```{r side by side comparison of M0 and M1}
grid.arrange(p_m0_1,p_m0_2,p_m1_2,,p2_m1, nrow = 2, ncol = 2)
grid.arrange(p_m0_3, p_m0_4, p_m1_3, p_m1_4, nrow = 2, ncol = 2)
```

Plot grid 1:
Observe in the bottom left: the mean residual line follows a slight sigmoid shape, 
suggesting the linear assumption is still violated for model 1.
On a positive note, the 'n' pattern about the mean is less pronounced for model 1.
The bottom right Q-Q plot looks slightly more aligned to the normal line at the lower tail (compared to M0).This suggests the prediction for the low count days is better.

Plot grid 2: 
Observe the leverage and thus influence is larger for the outliers of m1.

## Model M2: Nonlinear temperature
Observe from the temperature vs count scatter plot, there is some non linear relationship as we can see some curvature (blue line).
Hence, if we add the quadratic 'mean temperature' term to model m1, the underlying structure maybe better covered by the model:

```{r side by side comparison of M1 and M2}
grid.arrange(p_m1_1, p_m1_2, p_m2_1, p_m2_2, nrow = 2, ncol = 2)
grid.arrange(p_m1_3, p_m1_4, p_m2_3, p_m2_4, nrow = 2, ncol = 2)
```          

```{r thing}
grid.arrange(p1_m1,p2_m1,p1_m2,p2_m2, nrow = 2, ncol = 2)
           
```
M2 shows a general allignment with linear modelling assumptions - 
no big violations.
Observe that both models look almost identical from the diagnostic plots, which indicates further comparison may be undertaken using the scores.

M2 shows a general alignment with linear modelling assumptions - there are
no notable violations.
Observe that both models look almost identical from the diagnostic plots, 
which indicates further comparison may be undertaken using the scores.
We observe the residuals vs leverage plot improves for m2;
that is the leverage of the outliers decreases.



## Model M3: [your extension title]
The expression is

Rationale for M3:
To mediate the heavy right tailed QQ plots seen in (m0-2), we apply the log 
transform to the count response.

m3 <- lm(log(count+1) ~ temp_mean + I(temp_mean^2) + trend + 
               factor(month) + factor(dow)  + is_covid + is_holiday, data = cycle_daily_df)

```{r M3 compared to M2}
grid.arrange(p_m2_1, p_m2_2, p_m3_1, p_m3_2, nrow = 2, ncol = 2)
grid.arrange(p_m2_3, p_m2_4, p_m3_3, p_m3_4, nrow = 2, ncol = 2)
```

```{r improvements from adding is holiday variable to m3}

library(ggplot2)

# Visualization 1: Do holidays actually look different?
ggplot(cycle_daily_df, aes(x = factor(is_holiday), y = count)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Impact of Holidays on Daily Cycle Count",
       x = "Is Holiday (0=No, 1=Yes)", y = "Cycle Count")

# Visualization 2: The COVID 'Step'
ggplot(cycle_daily_df, aes(x = date, y = count, color = factor(is_covid))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(title = "Cycling Trends: Pre-Covid vs. Covid Era")


```


# Evaluation
## Statistical Performance

<!--
Required content:
  1. Table 1: LOYO CV scores for M0-M3.
  2. Table 2: CV RMSE and DS by month for your final model.
  3. Discussion: do the scores agree on which model is best? 
     If not, explain why and state your final model choice explicitly.
-->

Table of the Leave-one-year-out CV scores.

```{r Table 1 LOYO CV scores for M0-M3}

knitr::kable(table1_final, 
             digits = 2, 
             caption = "Table of the Leave-one-year-out CV scores")
```

Table of the RMSE and DS by month

```{r Table 2 of the RMSE and DS by month}

knitr::kable(table2_final, 
             digits = 2, 
             caption = "Table of the RMSE and DS by months")
```

## Practical Performance

<!--
One subsection per planning question (5.1-5.5 from the assignment brief).
Each must include mandatory fixed-format numbers and real-world interpretation.
-->

### 5.1 Temperature effects and nonlinearlity

The table shows the different score values for model 2, but using 3 different temperature variables:temp mean, temp min and temp max.

```{r M2 fit with temp_mean, temp_min, and temp_max}

knitr::kable(m2_comparison, 
             digits = 2,
             caption = "Model M2 fitted with daily temp mean, temp min, temp max.")
```

The model with temp max in the nonlinear term has the best predictive performance because it has the lowest value for each score.

At $5 ^\circ$ c,a $1 ^\circ$ c increase is associated with approximately 108 additional cyclists.

At $15 ^\circ$ c,a $1 ^\circ$ c increase is associated with approximately 166 additional cyclists.

We can see the effects a $1 ^\circ$ increase has in Spring and winter specifically. 

On a Spring day at $5 ^\circ$ c,a $1 ^\circ$ c increase is associated with approximately 343 additional cyclists.

On a Spring day at $15 ^\circ$ c,a $1 ^\circ$ c increase is associated with approximately 424 additional cyclists.

On a Winters day at $5 ^\circ$ c,a $1 ^\circ$ c increase is associated with approximately 56 additional cyclists.

On a Winters day at $15 ^\circ$ c,a $1 ^\circ$ c increase is associated with approximately 137 additional cyclists.

As we can see, there is huge variablity between seasons. For transport planning this implies that looking at the range of temperatures for the day is more importantas it will have a bigger effect. 


# Discussion and limitations

<!--
3-5 paragraphs covering:
  - Whether model assumptions appear satisfied.
  - Key limitations of the model and data.
  - What ideas would improve prediction.
  - Whether findings generalise to other cities or future years.
Do not introduce new results here.
-->

# Appendix - References
$^{1}$ https://mccyclingdiary.blogspot.com/2022/08/london-edinburg-london-lel-2022-edition.html

# Appendix - Code


All analysis code is in `code.R` and is reproduced below.

```{r print-code, echo=TRUE, eval=FALSE, code=readLines("code.R")}
```


