---
title: 'Project 2: Modelling daily cycling demand in Edinburgh'
author: 'Group members: Alice Miller, Saba Sitter, Jake Saunders'
date: 'Word count: [XXX words]'
output:
  html_document:
    df_print: paged
  pdf_document:
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
header-includes:
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{amsmath}
- \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,    # This hides all code from the main body 
  message   = FALSE, 
  warning   = FALSE
)
source("code.R") # Runs all models and creates tables/plots in the background
```

# Introduction

The "cycle daily" dataset  consists of approximately 2,180 daily observations ranging from 2020 to 2025, sourced from Edinburgh's network of automatic cycling counters, along with local weather and temporal data. This data comes from sensors at specific locations across the city, so it's possible not every cyclist is accounted for. However, they are a good indicator of how cycling demand varies across different days and weather patterns. Edinburgh relies heavily on public transport for its residents and tourists to navigate the city. For future funding allocation and planning, it is important to know how many cyclists are on the road in order to evaluate whether the infrastructure (cycle lanes, crossings etc) and investment in schemes (rentable Voi bikes) need to adapt to meet demand. The modelling goal is to develop a sequence of increasingly complex linear regression models to isolate how temporal and weather variables, such as the day of the week and the temperature, influence cycling demand. Ultimately, the goal is to extrapolate cycling demand to the future so that Edinburgh City Council and Transport Scotland can optimise infrastructure investment over the next decade to align with their active travel priorities.


# Exploratory data analysis

To begin investigating the impact of weather on the number of daily cyclists, we begun with some exploratory data analysis. Table 1 displays the summary statistics for the weather data and number of daily cyclists is as follows:

## Summary Statistics table for count, temp_mean, temp_min, temp_max

```{r summary stats exploratory table, echo=FALSE}

knitr::kable(summary_stats, 
             digits = 2, 
             caption = "Summary Statistics for count, temp mean, temp min, temp max")
```
We can see that the number of daily cyclists varies from $1,264$ to $43,864$, an increase of $\approx 3470 \%$, and the variation is very high. Intuitively, one would think that a warmer average weather would lead to more cyclists. 

We then plotted a time series of the daily cyclist count from 2020 to 2025. 

## Time series of daily counts 2020-2025 with a smoother.

```{r time series, fig.cap="Time series of daily counts 2020-2025 with a smoother"}
plot_timeseries
```


We can see from figure 1 that there are yearly fluctuations, with peaks and troughs occuring yearly. Overall, the cyclist count seems to be decreasing. There is a visible peak in 2022, which we investigate. We then created boxplots of the daily cyclist count by month and by the day of week. 


## Boxplots of count by month and by day of week.

```{r boxplots of count by month and day of week, fig.cap="Boxplots of count by month and by day of week"}
grid.arrange(plot_month,plot_dow, nrow = 2)
```


We can see from figure 2 that the winter months have slightly lower cyclist counts, with the other months remaining relatively constant. This indicates that the month does not have a tremendous impact on the number of daily cyclists outside of January and December. From the second boxplot, there appear to be less cyclists on Mondays and Sundays, with the other days again appearing relatively constant. 

## Scatter plot of count vs temp_mean.

```{r scatter plot of count vs temp_mean, fig.cap="Scatter plot of count vs temperature mean"}
plot_temp_scatter
```

Figure 3 shows a scatter plot of daily counts vs the mean temperature for that day, with the red line being the regression line. There is a general trend as the temperature increases, so does the daily count which is as expected.

## Extra Plots

```{r Boxplots of weekend and weekday, fig.cap="Boxplots of count vs weekend or weekday over the months (across 2020 to 2025)"}
plot_weekend_seas
```

Figure 4 shows for each month a boxplot for the different day types (weekends or weekdays). Generally the medians tend to be higher for the weekdays across all the months, but by varying amounts. June and Novemeber seems to have the biggest variance between the two days types

# Baseline model M0

<!--
Required content:
  1. State M0 mathematically using a numbered equation.
  2. Coefficient table (use kable, not raw output).
  3. Residuals vs fitted and normal QQ diagnostic plots, side by side.
  4. Brief assessment: what does M0 get right and wrong? 
     What motivates M1?
-->

The equation for M0 is:

$$\text{count}_i = \beta_0 + \beta_1 \text{temp\_mean}_i + \beta_2 \text{weekend}_i + \beta_3 \text{month}_i + \epsilon_i$$$$\epsilon_i \sim N(0, \sigma^2)$$

Table 2 shows the summary statistics for model 0.

```{r summary for m0, echo=FALSE} 
##notice the p values all 0 bc rounding

knitr::kable(summary(m0)$coefficients, 
             digits = 2, 
             caption = "Summary of model M")
```

Residuals vs fitted and normal QQ diagnostic plots.

```{r r vs f and qq plots m0, fig.cap="Boxplots of count by month and by day of week"}
grid.arrange(p1_m0,p2_m0, ncol = 2)
plot(m0, which = 4)

```

Model 0 (m0) shows a mean variance of 0, independent of the fitted values.
The residuals show a non-linear pattern, with an arguable 'n' shape above the mean line, indicating there is an underlying non-linear structure yet to be included in our model.

The QQ plot demonstrates right skewing and heavy tail in the positive quantiles - 
  thus the normality assumption for linear regression is violated.
This indicates the model is particularly bad at predicting high count observations.

# Observe the high influence points 953 823, 1806

```{r}
influential_rows <- cycle_daily_df[c(953, 823, 1806), ]
influential_rows
```

The highest influence date is potentially from the London, 
Edinburgh cycling festival and fringe combination.


# Proposed models M1--M3

<!--
For each model:
  - State the model mathematically.
  - Justify every new term BEFORE showing results.
  - Show diagnostic plots and comment on improvement over the previous model.
-->

## Model M1: Trend and factor 

###mathematical expression

Rational for investigating m1 over m0:
m1 has the granularity for the effect of different weekdays rather than mere weekend or non weekend effects.
Changing 'month' to a factor variable allows for building a seasonal profile rather than assuming the month index (1-12) is linear with the count.

We omit the weekend term and numerical version of month to avoid collinearity with their 'factor' versions and because the covariate dow is collinear with weekend.


```{r side by side comparison of M0 and M1}
grid.arrange(p1_m0,p2_m0,p1_m1,p2_m1, nrow = 2, ncol = 2)
```

Observe in the bottom left: the mean residual line follows a sigmoid shape, 
suggesting the linear assumption is still violated for model 1.
On a positive note, the 'n' pattern is less pronounced for model 1.
The bottom right Q-Q plot looks slightly more aligned to the normal line at the lower tail (compared to M0), which suggests the prediction for low count days is better.


## Model M2: Nonlinear temperature
Observe from the temperature vs count scatter plot, there is some non linear relationship as we can see some curvature (blue line).
Hence, if we add the quadratic 'mean temperature' term to model m1, the underlying structure maybe better covered by the model:


```{r}
grid.arrange(p1_m1,p2_m1,p1_m2,p2_m2, nrow = 2, ncol = 2)
           
```
M2 shows a general allignment with linear modelling assumptions - 
no big violations.
Observe that both models look almost identical from the diagnostic plots, which indicates further comparison may be undertaken using the scores.


## Model M3: [your extension title]

-----
IDEATION:
If your primary goal is to fix the plots you just showed me:

Choose the Log-transform to fix the Q-Q plot (the outliers).

Choose the Interaction or Temperature Range to fix the Residuals vs Fitted (the "wiggle").


------

# Evaluation
## Statistical Performance

<!--
Required content:
  1. Table 1: LOYO CV scores for M0-M3.
  2. Table 2: CV RMSE and DS by month for your final model.
  3. Discussion: do the scores agree on which model is best? 
     If not, explain why and state your final model choice explicitly.
-->

## Practical Performance

<!--
One subsection per planning question (5.1-5.5 from the assignment brief).
Each must include mandatory fixed-format numbers and real-world interpretation.
-->

# Discussion and limitations

<!--
3-5 paragraphs covering:
  - Whether model assumptions appear satisfied.
  - Key limitations of the model and data.
  - What ideas would improve prediction.
  - Whether findings generalise to other cities or future years.
Do not introduce new results here.
-->

# Appendix: Code

All analysis code is in `code.R` and is reproduced below.

```{r print-code, echo=TRUE, eval=FALSE, code=readLines("code.R")}
```


